---
title: "1. Installing R and Intro"
author: "Walter Mymala, Mihir Kulkarni"
output: html_document
---

# What is R?

* https://www.r-project.org/about.html

* https://skdeshpande91.github.io/wharton_moneyball/more_resources.html

Downloading R: 

R download mirrors: https://cran.r-project.org/mirrors.html

* Offical link to download coding language to your computer

RStudio download: https://www.rstudio.com/products/rstudio/download

<<<<<<< HEAD
* IDE (Integrated development environments) to work on R code. IDEs are useful for any coding language you are doing since it provides a global environment to keep track of your code, allows you to download packages, auto-fill code, and much more resources to keep you organized! Can also switch themes since default is light mode!

# Starting out

* Script Screen - The window that we are on right now. Allows you to type information and code! 

* Console - Window to type commands To remove data, click the broom button at the top of the window

* Environment | History | Tutorial - Window lists all data, values, functions you ran. To remove data, click the broom button at thetop of the window | Allows you to see code you recently ran code | Allows you to download packages to learn how to use R!

* Files | Plots | Packages | Help | Viewer - See files on your computer | Plots - Window where you can view graphs/plots you run in R | Packages - List of bundles code and documents and allows you to run other code that the basic R codes do not allow you to run. | Help - Database to see how to use a certain function or built in dataset that R has use a '?' and type in the function/dataset of inserest to trigger help window. Ex) ?mean


# R code chunks (using R Markdown)

* R Script is a notebook for code, R Markdown is where you can knit your documents (top left of window) to a different document type.

We can execute R code in the document with code chunks
```{r}
x <- 218 # Making variables: Equivalent to x = 218
x + 238 # you can also run code in console as well!
```

Here are the operators in R: https://www.statmethods.net/management/operators.html 

```{r}
a = 10 # Creates an object x with value 10
b = 9
a - b	# subtraction
a * b	# multiplication
a / b	# division
5 ^ 2 # or **	exponentiation
9 %% 6	# modulus (remainder value of division), 5 %% 2 is 1
6 %/% 4 # integer division (division answer w/o remainder), 8 %/% 3 is 2
```

Simple Vector
```{r}
x <- 6 # Overwrites original x value with value 6

y <- c(1,2,3,4) # c() function creates a vector of 4 values
length(y) #  tells us how many elements are in the vector, would output "4"
class(y) # tells us the type of element the object is, would output "numeric"
```

Built in functions: https://www.javatpoint.com/r-built-in-functions
```{r}
log(2)
factorial(5)
exp(log(7))
abs(-9)
```

Basic plots:
```{r}
D <- mtcars # loading in built-in dataset in R and setting it as variable D. R has a bunch of built-in ones
data() # check built-in data sets R has 
View(mtcars) # see which variables and dat is in mtcars

# Scatterplot 
plot(D$mpg, D$hp) # plot of mpg against hp(gross horsepower)
plot(D$mpg, D$hp, xlab = "Miles Per Gallon", ylab = "Horsepower", col = ifelse(D$mpg > 20, "#1b6618", "tomato3")) # all you need is your data to make basic plots, but adding additional info will help people visualize your values better
abline(v = 20.09062, col="purple") # adds straight lines to any plot. mean(D$hp) is 20.09062
legend("topright", legend = c("mpg > 20", "mpg < 20"), col=c("#1b6618", "tomato3"), pch = 1)

# Histogram 
hist(D$mpg, main = "Histogram of Motor Trend Car Road Tests mpg", xlab = "Miles Per Gallon")

# Boxplot
boxplot(D$mpg, main = "Histogram of Motor Trend Car Road Tests mpg", ylab = "Miles Per Gallon")

# Normal QQ Plots - determines if data is close to being normal distribution
qqnorm(D$mpg)
```

Working directory: "File path on your computer that sets the default location of any files you read into R" (Google). By default, the working directory for R your R file is the directory that contains the Rmd or Rscript document
```{r}
# setwd('/Users/___/Documents/____/____/____')
getwd() # prints directory of where the file is
```


# Importing/Reading in datasets. 
* Datasets are tables or places where information/data is stored for someone to do data analysis on. There are multiple types of datasets in R that can be read:

* This is the majority of how we will be using data in R, rather than just creating the datasets

* To import datasets, go to File -> Import Dataset -> Choose either From Excel or From Stata depending on the type of dataset you are using

* .RData : binary file written by R
* .csv
* .txt
* .xls and .xlsx (Excel)
* etc.

```{r}
# .csv and .txt files
# load() for .RData
# read.table() 
?read.table
# read.csv() a specific case of read.table()
?read.csv
```

```{r}
readxl::read_excel # reading in excel file

# .xls
library("readxl") # load library required to read excel, will touch on libraries later!
Excel <- read_excel("/Users/yeezywally_510/Documents/ARE 115A/KIDS Data.xlsx")
```











