---
title: "1. Installing R and Intro"
author: "Walter Mymala, Mihir Kulkarni"
output: html_document
---

# What is R?

* https://www.r-project.org/about.html

Example of you can do on R!:

* https://skdeshpande91.github.io/wharton_moneyball/more_resources.html

* https://github.com/WMymala/rNBA/blob/main/2.%20FT%20Comps.Rmd

Downloading R: 

R download mirrors: https://cran.r-project.org/mirrors.html

* Offical link to download coding language to your computer

RStudio download: https://www.rstudio.com/products/rstudio/download

* IDE (Integrated development environments) to work on R code. IDEs are useful for any coding language you are doing since it provides a global environment to keep track of your code, allows you to download packages, auto-fill code, and much more resources to keep you organized! Can also switch themes since default is light mode!

# Starting out

* Script Screen - The window that we are on right now. Allows you to type information and code! 

* Console - Window to type commands To remove data, click the broom button at the top of the window

* Environment | History | Tutorial - Window lists all data, values, functions you ran. To remove data, click the broom button at thetop of the window | Allows you to see code you recently ran code | Allows you to download packages to learn how to use R!

* Files | Plots | Packages | Help | Viewer - See files on your computer | Plots - Window where you can view graphs/plots you run in R | Packages - List of bundles code and documents and allows you to run other code that the basic R codes do not allow you to run. | Help - Database to see how to use a certain function or built in dataset that R has use a '?' and type in the function/dataset of inserest to trigger help window. Ex) ?mean


# R code chunks (using R Markdown)

* R Script is a notebook for code, R Markdown is where you can knit your documents (top left of window) to a different document type.

We can execute R code in the document with code chunks
```{r}
x <- 218 # Making variables: Equivalent to x = 5
x + 238 # you can also run code in console as well!
```

Here are the operators in R: https://www.statmethods.net/management/operators.html 

```{r}
a = 10
b = 9
a - b	# subtraction
a * b	# multiplication
a / b	# division
5 ^ 2 # or **	exponentiation
9 %% 6	# modulus (remainder value of division), 5 %% 2 is 1
6 %/% 4 # integer division (division answer w/o remainder), 8 %/% 3 is 2
```

Built in functions: https://www.javatpoint.com/r-built-in-functions
```{r}
log(2)
factorial(5)
exp(log(7))
abs(-9)
```

Working directory: "File path on your computer that sets the default location of any files you read into R" (Google). By default, the working directory for R your R file is the directory that contains the Rmd or Rscript document
```{r}
# setwd('/Users/nhy163/Documents/Alex/Teaching/first_project')
getwd() # prints directory of where the file is
```


Reading in datasets. Datasets are tables or places where information/data is stored for someone to do data analysis on. There are multiple types of datasets in R that can be read:
 
* .RData : binary file written by R
* .csv
* .txt
* .xls and .xlsx (Excel)
* etc.

```{r}
# .csv and .txt files
load() # for .RData
read.csv() # a specific case of read.table()
```

```{r}
readxl::read_excel # reading in excel file

# .xls
library("readxl") # load library required to read excel, will touch on libraries later!
Excel <- read_excel("/Users/yeezywally_510/Documents/ARE 115A/KIDS Data.xlsx")
```

# More Resources:

* R for data science textbook - https://r4ds.had.co.nz/ 

* Exploratory Data Analysis with R - https://bookdown.org/rdpeng/exdata/ 

* Most important: For projects and code in general, Google and StackOverflow are your best friend! Web scraping data for projects is great too!







