---
title: "2.-Analyzing-Baseball-Data-Lahman_Database"
author: "Walter Mymala, Mihir Kulkarni"
output: html_document
---

Book: Analyzing Baseball Data with R (Second Edition) - Chapman and Hall

Link for the book if you really want to read it (totally optional): https://libgen.is/book/index.php?md5=AA1AE52B2BA2B01497FACAD2B7E4807F

# Intro

Lahman is a database w/ pitching, hitting, and fielding statistics for the entire history of professional baseball from 1871 to the current. csvs in Lahman on pp. 4 Analyzing Baseball Data.

# Chapter 2 of Book 

# 2.3 Tidyverse

```{r}
## install.packages("Lahman"); Install a library. Libraries have additional fxns and tricks that regular R cannot run
library(Lahman) # runs librar(ies) in file
library(tidyverse) # important library for data analysis. dplyr good for extracting data and ggplot good for visualizing observations
```

```{r}
# piping fxn (%>%) allows us to write neater code for nested functions
Master %>% head(1) # biographic info. head(Master, 1) without using piping fxn
Batting %>% head(5) # .csv with seasonal batting stats. Try to run Pitching .csv and Fielding .csv on your own
Teams %>% head(5) # seasonal data at the team level going back to 1871.
```

# 2.4 Data Frames

```{r}
spahn <- read.csv("/Users/yeezywally_510/Desktop/MoneyBall/spahn.csv") # read dataset
#View(spahn_data)
class(spahn) # dataframe (df) - rectangular table of data, rows usually correspond to different observations, and columns of the table correspond to different variables collected on the observations.
```

dplyr functions: https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html
```{r}
spahn %>% slice(1:3) %>% select(1:10) # == select(slice(spahn, 1:3), 1:10); shows first three rows and columns 1 though 10 of spahn df
spahn %>% slice(1:10) %>% select(Age, W, L, ERA) # custom select columns 
```

```{r}
# get 5 number summary stats for Spahn's ERA
spahn %>%
  summarize(LO = min(ERA),
            QL = quantile(ERA, .25),
            M = median(ERA), QU = quantile(ERA, .75), HI = max(ERA))

# Find age when Spahn had lowest ERA
spahn %>% filter(ERA == min(ERA)) %>% select(Age)
```

$$FIP = {13HR+3BB-2K}/IP$$ - FIP (Field Independent Pitching) scales ERA. "Measurement of a pitcher’s performance that strips out the role of defense, luck, and sequencing, making it a more stable indicator of how a pitcher actually performed over a given period of time than a runs allowed based statistic that would be highly dependent on the quality of defense played behind him" (Fangraphs). It's also a future ERA estimator for pitchers.
```{r}
# manipulations
spahn %>%
  mutate(FIP = (13 * HR + 3 * BB - 2 * SO) / IP) -> spahn # add new column variable to df with mutate()

# see which years Spahn had lowest FIP
spahn %>%
  arrange(FIP) %>%
  select(Year, Age, W, L, ERA, FIP) %>%
  head() # his best FIPs were during the middle of his career

# mutate TOT (multiple teams) value from Tm (Team) column so only 2 unique values (factors) in the column 
# spahn <- spahn %>%
#  mutate(Tm = factor(Tm, levels = c("BSN", "MLN")))
 
# compare Spahns stats when he played in Boston vs. in Milwaukee
 spahn %>%
  group_by(Tm) %>%
  summarize(mean_W.L = mean(W.L, na.rm = TRUE),
            mean_ERA = mean(ERA),
            mean_WHIP = mean(WHIP),
            mean_FIP = mean(FIP))
```

# 2.5 Vector Functions

```{r}
# Wins and Losses from 1946 to 1952, his years in Boston
W <- c(8, 21, 15, 21, 21, 22, 14)
L <- c(5, 10, 12, 14, 17, 14, 19)

Win.Pct <- 100 * W / (W + L) # vector combining 2 vectors above

Year <- seq(from = 1946, to = 1952) # seq() runs patterned data; 1946 : 1952 is also valid

Age <- Year - 1921 # Spahn was born in 1921

plot(Age, Win.Pct) 
```

```{r}
# Vector Index
W[c(1, 2, 5)] # extract portions of vectors
W[1:4]
W[-c(1, 6)] # w/ minus index, we remove vector indicies

# Logical variable, either have >, < or =
Win.Pct > 60 # prints TRUE or FALSE
Win.Pct == max(Win.Pct)
Year[W + L > 30]
```

# 2.6 Objects and Containers in R

String variables like the team names and players are stored as that are represented by letters enclosed by double quotes (characters) and numbers (integers). 

As an example, we wish to explore information about the World Series in the years 2008 through 2017
```{r}
# 5 new vectors, 
Year <- 2008 : 2017 
NL <- c("PHI", "PHI", "SFN", "SLN", "SFN",
        "SLN", "SFN", "NYN", "CHN", "LAN")
AL <- c("TBA", "NYA", "TEX", "TEX", "DET",
        "BOS", "KCA", "KCA", "CLE", "HOU")
Winner <- c("NL", "AL", "NL", "NL", "NL",
            "AL", "NL", "AL", "NL", "AL")
N_Games <- c(5, 6, 5, 7, 4, 7, 7, 5, 7, 7)

# Tibble and dataframes are both 2D data structures. Tibble is a dataframe that makes using tidyverse easier and loads data faster than df
WS_results <- tibble(
  Year = Year, NL_Team = NL, AL_Team = AL,
  N_Games = N_Games, Winner = Winner)
WS_results
```

```{r}
# # grep() matches patterns in the values
grep("NY", c(AL, NL), value = TRUE) 

# summarize() will create a new data frame with group_by() which will group the data frame for each World Series league winner (variable Winner). We count the rows by use of the n() function.
WS_results %>%
  group_by(Winner) %>%
  summarize(N = n()) -> WS
WS

# bar graph showing frequencies of code directly above
ggplot(WS, aes(x = Winner, y = N)) +
  geom_col()
```

```{r}
# factors - special way of representing character (categorical) data
WS_results %>%
  group_by(NL_Team) %>%
  summarize(N = n())

# Similar to lines 123-126 but now we redefine the NL_Team variable by making NL_Team values into factors. Factors help the data not be put into alphabetical order and can omit values that are not part of the levels into an NA value
WS_results <- WS_results %>%
  mutate(NL_Team = factor(NL_Team,
                          levels = c("NYN", "PHI", "CHN",
                                     "SLN", "LAN", "SFN")))
WS_results

# similar to lines 146-148 but now teams listed in order specified by factor()
WS_results %>%
  group_by(NL_Team) %>%
  summarize(N = n())
```

```{r}
# Lists - object in R that allows you to store elements of different types like integers, characters, even other vectors 
world_series <- list(Winner = Winner, Number.Games = N_Games,
                     Seasons = "2008 to 2017")
world_series

# accessing elements in a list
world_series$Number.Games # same as world_series[[2]] or world_series["Number.Games"]
world_series$Number.Games[10] # access last element of Number.Games
```

# R functions - used to implement implement specific computations or graphs of interest that are not in R libraries or base R
```{r}
# Function to compute a player’s home run rates for a collection of seasons; returns a list. One inputs a vector age of player ages, a vector hr of home run counts, and a vector ab of at-bats
hr_rates <- function(age, hr, ab) {
  rates <- round(100 * hr / ab, 1) # rounds hr rates to one digit
  list(x = age, y = rates) # fxns return last line of code in itself unless return command is used
}

# data for Mickey Mantle's(NY Yankees legend) seasons from 1951 to 1961
HR <- c(13, 23, 21, 27, 37, 52, 34, 42, 31, 40, 54)
AB <- c(341, 549, 461, 543, 517, 533, 474, 519, 541, 527, 514)
Age <- 19 : 29
Mantle_HRrates <- hr_rates(Age, HR, AB)
names(Mantle_HRrates) <- c("Age", "HR_rates") # rename list elements
Mantle_HRrates

plot(hr_rates(Age, HR, AB)) # Mantle’s home run rates rose steadily in the first six seasons
```

# 2.10 Splitting, Applying, and Combining Data
```{r}
# We want to answer: “Who hit the most home runs in the 1960s?”. We can filter out data to only include the 1960s
Batting %>%
  filter(yearID >= 1960, yearID <= 1969) ->  Batting_60 

# split the data frame by each player id, and then compute the sum of home runs for each player
Batting_60 %>%
  group_by(playerID) %>%
  summarize(HR = sum(HR)) -> hr_60 # Batting_60 now computes total number of home runs for each player in the data frame

# sort data so best HR hitters are on top
hr_60 %>%
  arrange(desc(HR)) -> hr_60
head(hr_60, 4) # Harmon Killebrew, Hank Aaron, Willie Mays, and Frank Robinson were the top HR hitters in the 1960s
```

What's the big picture?:  We have dataset(s) to work with. We can use the tidyverse data to pull out specific data that we have questions on like: "If Shohei Ohtani were to focus on only pitching or batting, which one should he choose/be more effective in?" We can also do data visualization with ggplot to support your claim.


