---
title: "3.-Analyzing-Baseball-Data-Graphics"
author: "Walter Mymala, Mihir Kulkarni"
output: html_document
---

Book: Analyzing Baseball Data with R (Second Edition) - Chapman and Hall

Download Files for code at: https://github.com/beanumber/baseball_R > Code (Green Button) > Download

Files used: 

* hofbatting.csv

* fields.csv

* all1998.csv

# Chapter 3

```{r}
library(tidyverse)

# Stats of HOF players excluding pitchers
hof <- read_csv("/Users/yeezywally_510/Desktop/MoneyBall/baseball_R-master/data/hofbatting.csv")
```

HOFs played in different baseball eras. We split player's mid-career to see which era they fall under. There are 7 different eras the book defines:

* “19th Century” (up to the 1900 season), “Dead Ball” (1901 through 1919), “Lively Ball” (1920 though 1941), “Integration” (1942 through 1960), “Expansion” (1961 through 1976), “Free Agency” (1977 through 1993), and “Long Ball” (after 1993)

```{r}
# creates 2 new variables (MidCareer and Era) to classify which era each hof batter played for
hof <- hof %>%
  mutate(MidCareer = (From + To) / 2,
         Era = cut(MidCareer,
                   breaks = c(1800, 1900, 1919, 1941,
                              1960, 1976, 1993, 2050),
                   labels = c("19th Century", "Dead Ball",
                              "Lively Ball", "Integration",
                              "Expansion", "Free Agency",
                              "Long Ball")))

hof_eras <- summarize(group_by(hof, Era), N = n())
hof_eras # see how many players played in which eras
```

```{r}
# Graphs for categorical variables - Bar Graph & Dot Plot

# aes() function defines aesthetics & geom_bar() makes a bar graph
ggplot(hof, aes(x = Era)) + geom_bar()

# adding axes labels and title
ggplot(hof, aes(Era)) +
  geom_bar() +
  xlab("Baseball Era") +
  ylab("Frequency") +
  ggtitle("Era of the Nonpitching Hall of Famers") 

# Dot Plot - use geom_point() to display alternative graphical displays for a table of frequencies of a character variable
ggplot(hof_eras, aes(Era, N)) +
  geom_point() +
  xlab("Baseball Era") +
  ylab("Frequency") +
  ggtitle("Era of the Nonpitching Hall of Famers") +
  coord_flip()
```

OPS - Measures the ability to get on base. Divide the total number of hits, walks, and times hits by a pitch by the number of times at bat plus walks. 
```{r}
# Graphs for categorical variables - 1D scatterplot & Histogram

# 1D scatterplot (different from scatterplot!)- Use geom_jitter() for a number line graph, where the values of the statistics are plotted over a number line ranging over all possible values of the variable.
ggplot(hof, aes(x = OPS, y = 1)) +
  geom_jitter(height = 0.6) + ylim(-1, 3) +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  coord_fixed(ratio = 0.03) # In the df hof, the OPS is mapped to the x aesthetic and the dummy variable y is set to a constant value. The theme() elements are chosen to remove the tick marks, text, and title from the y-axis.

# histogram - Use geom_histogram(), where values are grouped into bins of equal width and bin frequencies are non-overlapping bars over the bins
ggplot(hof, aes(x = OPS)) +
   geom_histogram()

# An issue w/ histograms is of bin #s, but geom_histogram() will usually make good choices for the bins to display good data distribution. However, you can customize bin #s in geom_histogram() w/ argument "breaks". Lines of the bars are colored blue and the bars are filled in white
ggplot(hof, aes(x = OPS)) +
  geom_histogram(breaks = seq(0.4, 1.2, by = 0.1),
                 color = "blue", fill = "white")
```

# 3.5 Two Numeric Variables

Scatterplot is for graphing two numeric variables. geom_point() makes the plot and geom_smooth() shows association btw the 2 variables. We now compare MidCareer and OPS for better visualization.
```{r}
# scatterplot
ggplot(hof, aes(MidCareer, OPS)) +
  geom_point() +
  geom_smooth()

library(ggrepel) # for geom_text_repel() to see outliers
ggplot(hof, aes(MidCareer, OPS)) +
  geom_point() +
  geom_smooth() +
  geom_text_repel(data = filter(hof, OPS > 1.05 | OPS < .5),
             aes(MidCareer, OPS, label = Player))
```

```{r}
# Building a graph, step-by-step (OPS - the ability to get on base, and SLG - ability to advance runners already on base)
(p <- ggplot(hof, aes(OBP, SLG)) + geom_point()) #  The outlier in the bottom-left makes points fall in a small region of the plot

# xlim() and ylim() functions change the limits of horizontal and vertical axes. New choices of the limits, (0.25, 0.50) for the horizontal and (0.28, 0.75) for the vertical, remove the outlier and allow for more space in the upper-left section of the graph for labels
(p <- p +
   xlim(0.25, 0.50) + ylim(0.28, 0.75) +
   xlab("On Base Percentage") +
   ylab("Slugging Percentage"))

# or we could use scale_x_continuous() and scale_y_continuous() to do the same thing
# p <- p +
#   scale_x_continuous("On Base Percentage",
#                      limits = c(0.25, 0.50)) +
#   scale_y_continuous("Slugging Percentage",
#                      limits = c(0.28, 0.75))
```

A good measure of batting performance is the OPS statistic defined by OPS = OBP + SLG
```{r}
# To evaluate hitters in our graph on the basis of OPS, we draw constant values of OPS on the graph. We wish to draw a line where OPS = 0.7 to OPS = 1
(p <- p + geom_abline(slope = -1,
                      intercept = seq(0.7, 1, by = 0.1)))

# label the lines; the three arguments in annotate() are the x location and y location where the text is to be drawn, and label is the vector of strings of text to be displayed
p + annotate("text",
             x = rep(.27, 4) , y = c(.42, .52, .62, .72),
             label = paste("OPS = ",
                     c(0.7, 0.8, 0.9, 1.0)))
```

# A Numeric Variable and a Factor (Categorical) Variable
We usually compare the distributions of the numeric variable across different values of the factor. geom_jitter() constructs parallel boxplots across the factor
```{r}
# Suppose we want to know the home run rates defined by HR / AB and how they change over different eras for our Hall of Fame players

# add a new variable hr_rate to the data frame
hof <- mutate(hof, hr_rate = HR / AB)

# Parallel stripcharts; height = 0.1 to reduce amount of vertical jitter of the points
ggplot(hof, aes(hr_rate, Era)) +
  geom_jitter(height = 0.1)

# Parallel boxplots; coord_flip() flips the axes and displays the boxplots horizontally
 ggplot(hof, aes(Era, hr_rate)) +
  geom_boxplot() + coord_flip()
```

# 3.7 Comparing Ruth, Aaron, Bonds, and A-Rod

Write a new function get_birthyear() to get this information for an arbitrary player of interest. In MLB, a player’s age for a season is defined to be his age on June 30. To adjust a player’s birth year to see if his birthday falls in the first six months or not, adjusted birth year is stored in the variable birthyear. 
```{r}
library(Lahman)

# We will extract the player id and the birth year for a particular player. 

# The ifelse() function is useful for assignments based on a condition; if birthMonth >= 7isTRUE,thenbirthyear <- birthYear + 1,otherwisebirthyear <- birthyear
get_birthyear <- function(Name) {
  Names <- unlist(strsplit(Name, " "))
  Master %>%
    filter(nameFirst == Names[1],
           nameLast == Names[2])  %>%
    mutate(birthyear = ifelse(birthMonth >= 7,
                              birthYear + 1, birthYear),
           Player = paste(nameFirst, nameLast)) %>%
    select(playerID, Player, birthyear)
}

# get_birthyear() gets information for the players of interest. bind_rows() combines the players’ info into a single df, PlayerInfo.
PlayerInfo <- bind_rows(get_birthyear("Babe Ruth"),
                        get_birthyear("Hank Aaron"),
                        get_birthyear("Barry Bonds"),
                        get_birthyear("Alex Rodriguez")
               )

# Now that we have player id and birth years, create data frames for each of these four players. inner_join() gets the batting and age data, to match the rows of the batting data to those corresponding in the PlayerInfo df where playerID is equal. We create a new variable Age defined to be the season year minus the player’s birth year. New variable CHR contains the cumulative count of home runs for each player each season
Batting %>%
  inner_join(PlayerInfo, by = "playerID") %>%
  mutate(Age = yearID - birthyear) %>%
  select(Player, Age, HR) %>%
  group_by(Player) %>%
  mutate(CHR = cumsum(HR)) -> HRdata

# plot  
ggplot(HRdata, aes(x = Age, y = CHR, linetype = Player)) +
  geom_line()
```

# The 1998 Home Run Race

We will graphically view the famous home run duel between Mark McGwire and Sammy Sosa
```{r}
# all1998.csv contains all the play-by-play data for the 1998 season. fields.csv contains the names of all variables in the play-by-play file
fields <- read_csv("/Users/yeezywally_510/Desktop/MoneyBall/baseball_R-master/data/fields.csv")
data1998 <- read_csv("/Users/yeezywally_510/Desktop/MoneyBall/baseball_R-master/data/all1998.csv",
                     col_names = pull(fields, Header))

# BAT_ID gives the identification code for the player who is batting. filter() allows us to find id code where nameFirst = "Sammy" and nameLast = "Sosa" and same thing for McGuire
sosa_id <- Master %>%
  filter(nameFirst == "Sammy", nameLast == "Sosa") %>%
  pull(retroID)
mac_id <- Master %>%
  filter(nameFirst == "Mark", nameLast == "McGwire") %>%
  pull(retroID)

# Now that we have the player id codes, we extract McGwire’s and Sosa’s plate appearance data from the play-by-play data frame data1998
 hr_race <- data1998 %>%
  filter(BAT_ID %in% c(sosa_id, mac_id))
```

For each player, we are interested in collecting the current number of home runs hit for each plate appearance and graphing the date against the home run count. For each player, the two important variables are the date and the home run count. We write a new function cum_hr() that will extract these two variables given a player’s play-by-play batting data
```{r}
# GAME_ID identifies the game location and date (e.g. ARI199805110 indicates this particular play occurred at the game played in Arizona on May 11, 1998). We select the 4th through 11th characters of this string variable and assign this date to the variable Date w/ str_sub(). ymd() function converts the date to the more readable “year-month-day” format. arrange() sorts play-by- play data from the beginning to end of the season. EVENT_CD contains the outcome of the batting play (EVENT_CD value of 23 indicates that a HR has been hit). We define a new variable HR to be either 1 or 0 depending if a home run occurred, and the new variable cumHR records the cumulative number of home runs hit in the season using the cumsum(). The output of the function is a new data frame containing each date and the cumulative number of home runs to date for all plate appearances during the season
library(lubridate)
cum_hr <- function(d) {
  d %>%
    mutate(Date = ymd(str_sub(GAME_ID, 4, 11))) %>%
    arrange(Date) %>%
    mutate(HR = ifelse(EVENT_CD == 23, 1, 0),
           cumHR = cumsum(HR)) %>%
    select(Date, cumHR)
}

# map_df() iterates cum_hr() twice, once on Sosa’s batting data and once on McGwire’s batting data
hr_ytd <- hr_race %>%
  split(pull(., BAT_ID)) %>%
  map_df(cum_hr, .id = "BAT_ID") %>%
  inner_join(Master, by = c("BAT_ID" = "retroID"))
```

Graph of the 1998 home run race between Sammy Sosa and Mark McGwire.
```{r}
# graph - geom_line() constructs a graph of the cumulative home run count against the date. geom_hline() adds a horizontal line at the home run value of 62 and annotate() is applied to place the text string “62” above this plotted line
ggplot(hr_ytd, aes(Date, cumHR, linetype = nameLast)) +
  geom_line() +
  geom_hline(yintercept = 62, color = "blue") +
  annotate("text", ymd("1998-04-15"), 65,
           label = "62", color = "blue") +
  ylab("Home Runs in the Season")
```

What's the big picture?: We can do HELLA visualization with ggplot.

* For categorical variables: 1D scatterplots & Histograms

* For numeric variables: Scatterplots

* For categorical variables: Bar Graphs (parallel bar graphs), Dot Plots, Parallel stripcharts

